RationalGRL: A Framework for Argumentation and Goal Modeling

### 2017-06-27 introduction.tex
many improvements (see email)

Requirements Engineering (RE) is an approach to assess the role of a future information system within a human or automated environment. An important goal in RE is to produce a consistent and comprehensive set of system requirements covering different aspects of the system, such as general functional requirements, operational environment constraints, and so-called non-functional requirements such as security and performance.

One of the first activities in RE are the ``early-phase'' requirements engineering activities, which include those that consider how the intended system should meet organizational goals, why it is needed, what alternatives may exist, what implications of the alternatives are for different stakeholders, and how the interests and concerns of stakeholders might be addressed~\cite{yu1997towards}. This is generally referred to as goal modeling. Given the number of currently established RE methods using goal models in the early stage of requirements analysis (e.g.,~\cite{liu2004designing,donzelli2004goal,dardenne1993goal,chung2012non,castro2002towards}, overviews can be found in~\cite{van2001goal,kavakliL05}), there is a general consensus that goal models are useful in RE. Several goal modeling languages have been developed in the last two decades. The most popular ones include $i*$~\cite{Yu:1997:TMR:827255.827807}, Keep All Objects Satisfied (KAOS)~\cite{van2008requirements}, the NFR framework~\cite{chung2012non}, \textsc{Tropos}~\cite{giorgini2005goal}, the Business Intelligence Model (BIM)~\cite{horkoff2014strategic}, and the Goal-oriented Requirements Language (GRL)~\cite{Amyot:2010:EGM:1841349.1841356}.

A goal model is often the result of a discussion process between a group of stakeholders. For small-sized systems, goal models are usually constructed in a short amount of time, involving stakeholders with a similar background. Therefore, it is often not necessary to record all of the details of the discussion process that led to the final goal model. 
However, most real-world information systems -- e.g., air-traffic management, industrial production processes, or government and healthcare services -- are complex and are not constructed in a short amount of time, but rather over the course of several workshops. In such situations, failing to record the discussions underlying a goal model in a structured manner may harm the success of the RE phase of system development for several reasons.

\todo{Marc}{Sepideh}{Please improve 2 and 4 by adding references}

\begin{enumerate}
\item
It is well-known that stakeholders' preferences are rarely absolute, relevant, stable, or consistent~\cite{march1978bounded}. Therefore, it is possible that a stakeholder changes his or her opinion about a modeling decision in between two goal modeling sessions, which may require revisions of the goal model. If previous preferences and opinions are not stored explicitly, it is not possible to remind stakeholders of their previous opinions, thus risking unnecessary discussions and revisions. As the number of participants increases, revising the goal model based on changing preferences can take up a significant amount of time. \todo{Floris}{all}{Depending on whether we use preferences in argumentation models, we might want to rephrase this}
\item
Other stakeholders, who were not the original authors of the goal model, may have to make sense of the goal model, for instance, to use it as an input in a later RE stage. If these user have no knowledge of the underlying rationale of the goal model, it may not only be more difficult to understand the model, but they may also end up having the same discussions as the previous group of stakeholders.
\item
Alternative different ideas and opposing views that could potentially have led to different goal diagrams are lost. For instance, a group of stakeholders specifying a goal model for a user interface may decide to reduce softgoals ``easy to use'' and ``fast'' to one softgoal ``easy to use''. Thus, the resulting goal model will merely contain the softgoal ``easy to use'', but the discussion as well as the decision to reject ``fast'' are lost. This leads to a poor understanding of the problem and solution domain. In fact, empirical data suggest that this is an important reason of RE project failure~\cite{curtis1988field}. 
\item
It is not possible to reason about changing beliefs and opinions, and their effect on the goal model. A stakeholder may change his or her opinion, but it is not always directly clear what its effect is on the goal model. Similarly, a part of the goal model may change, but it is not possible to reason about the consistency of this new goal model with the underlying beliefs and arguments. This becomes more problematic if the participants constructing the goal model change, since modeling decisions made by one group of stakeholders may conflict with the underlying beliefs put forward by another group of stakeholders.
\end{enumerate}

The aim of this research is to resolve the above issues by designing a framework, which combines a methodology for the goal modeling process \todo{Floris}{Sepideh}{Do you know of any generic methodologies or heuristics for goal modeling (akin to Juretas GAM)?} with ideas from Artificial Intelligence on practical reasoning and argumentation \cite{atkinson2007} and simple software tool support. We identified several important requirements for our framework: (1) it must be able to formally model parts of the discussion process in the early-requirements phase of an information system; (2) it must be able to generate formal goal models based on the discussions; (3) it should have formal traceability links between goal elements and arguments in the discussions; (4) it must have tool support; (5) a methodology must be identified that allows the framework to be used by practitioners; and (6) the framework must identify arguments and rationales that might not have been found, or might have been lost, otherwise \todo{Floris}{all}{I think (6) should be replaced with something we actually do in the paper: we don't show that arguments etc. allows us to find rationales that would otherwise not have been found -- for this we would need an experiment. Tailor it towards our empirical work: ``the framework should be close to on the actual discussions stakeholders or designers have in the early requirements engineering phase''}.

In this context, the main research question is: \emph{What is a suitable framework to formally capture the discussions between stakeholders such that it can generate goal models, and how to implement this framework?} The first five requirements are the success criteria of our approach, that is, the satisfaction of the five requirements will result in a positive answer to the research question. \todo{Sepideh}{all}{The last sentence has to be changed in the end to reflect the result when the paper/evaluation is done. If we write it this way, it means we are unsure of the result.}

\subsection{Contributions}

The contributions of this paper are aligned with our five success criteria. In order to formalize discussions of the early requirements phase of an information system (requirement 1), we use a technique from argumentation and discourse modelling called \emph{argument schemes} \cite{walton-etal2004}. This allows us to formulate arguments and counter-arguments using so-called \emph{critical questions}. By combining arguments and critical questions, one can create a structured representation of a discussion. We first propose a set of argument schemes and critical questions we expect to find in the discussions that take place during the goal modelling process. We then validate this initial set with transcripts containing discussions about the architecture of an information system. This then gives us the final set of arguments and critical questions.

In order to generate goal models based on formalized discussions (requirement 2), we first formalize the list of arguments from requirement 1 in an argumentation framework. We formalize the critical questions as algorithms modifying the argumentation framework. We use argumentation techniques from AI in order to determine which arguments are accepted and which are rejected. We propose an algorithm to generate a GRL model based on accepted arguments. This creates traceability links from GRL elements to underlying arguments (requirement 3).

We implement our framework into an online tool called RationalGRL (requirement 4). The tool is implemented using Javascript. It contains  two parts, goal modeling and argumentation. The goal modeling part is a simplified version of GRL, leaving out features such as evaluation algorithms and key performance indicators . The argumentation part is new, and we develop a modeling language for the arguments and critical questions. The created GRL models in RationalGRL can be exported to jUCMNav~\cite{}, the Eclipse-based tool for GRL modeling, for further evaluation and  analysis.

Our final contribution is a methodology on how to develop goal models that are linked to underlying discussions. The methodology consists of two parts, namely argumentation and goal modeling. In the argumentation part, one puts forward arguments and counter-arguments by applying critical questions. When switching to the goal modeling part, the accepted arguments are used to create a goal model. In the goal modeling part, one simply modifies goal models, which may have an effect on the underlying arguments. This might mean that the underlying arguments are no longer consistent with the goal models.

@@

The first two sections are introductory: Section~\ref{sect:background} contains background and introduces the Goal-oriented Requirements Language (GRL)~\cite{} and argument schemes, while Section~\ref{sect:preview} provides a brief and high-level overview of our framework and methodology using an example.

The rest of the article is in line with the main contributions. We develop the list of argument schemes in two iterations, corresponding to two sections: In Section~\ref{sect:gmas}, we introduce an initial list of Argument Schemes for Goal Modeling (GMAS) and associated critical questions based on an existing argument scheme. In Section~\ref{sect:gmas:2}, we evaluate the initial list of argument schemes by annotating transcripts from discussions about an information system, which results in our final list of argument schemes and critical questions. In Section~\ref{sect}, we develop a formal model for argument schemes and critical questions and we describe RationalGRL tool in Section~\ref{sect:implementation}. In Section~\ref{sect:methodology}, we propose our methodology while we discuss the related work in Section~\ref{sect:relatedwork}.

>>>>

\subsection{Requirements Engineering and Goal Modeling}

Requirements Engineering (RE) is an approach to assess the role of a future information system within a human or automated environment. An important goal in RE is to produce a consistent and comprehensive set of requirements covering different aspects of the system, such as general functional requirements, operational environment constraints, and so-called non-functional requirements such as security and performance.

One of the first activities in RE are the ``early-phase'' requirements engineering activities, which include those that consider how the intended system should meet organizational goals, why it is needed, what alternatives may exist, what implications of the alternatives are for different stakeholders, and how the interests and concerns of stakeholders might be addressed~\cite{yu1997towards}. These activities fall under the umbrella of goal modeling. The are a large number of established RE methods using goal models in the early stage of requirements analysis (e.g.,~\cite{liu2004designing,donzelli2004goal,dardenne1993goal,chung2012non,castro2002towards}, overviews can be found in~\cite{van2001goal,kavakliL05}). Several goal modeling languages have been developed in the last two decades as well. The most popular ones include $i*$~\cite{Yu:1997:TMR:827255.827807}, Keep All Objects Satisfied (KAOS)~\cite{van2008requirements}, the NFR framework~\cite{chung2012non}, \textsc{Tropos}~\cite{giorgini2005goal}, the Business Intelligence Model (BIM)~\cite{horkoff2014strategic}, and the Goal-oriented Requirements Language (GRL)~\cite{Amyot:2010:EGM:1841349.1841356}.

\subsection{Problem Statement}

A goal model is often the result of a discussion process between a group of stakeholders. For small-sized systems, goal models are usually constructed in a short amount of time, involving stakeholders with a similar background. Therefore, it is often not necessary to record all of the details of the discussion process that led to the final goal model. However, most real-world information systems -- e.g., air-traffic management, industrial production processes, or government and healthcare services -- are complex and are not constructed in a short amount of time, but rather over the course of several workshops. In such situations, failing to record the discussions underlying a goal model in a structured manner may harm the success of the RE phase of system development for various reasons.

First, it is well-known that stakeholders' preferences are rarely absolute, relevant, stable, or consistent~\cite{march1978bounded}. Therefore, it is possible that a stakeholder changes his or her opinion about a modeling decision in between two goal modeling sessions, which may require revisions of the goal model. If previous preferences and opinions are not stored explicitly, it is not possible to remind stakeholders of their previous opinions, thus risking unnecessary discussions and revisions. As the number of participants increases, revising the goal model based on changing preferences can take up a significant amount of time.

Secondly, other stakeholders, such as new developers on the team who were not the original authors of the goal model, may have to make sense of the goal model, for instance, to use it as an input in a later RE stage or at the design and development phase. If these user have no knowledge of the underlying rationale of the goal model, it may not only be more difficult to understand the model, but they may also end up having the same discussions as the previous group of stakeholders.

Thirdly, alternative different ideas and opposing views that could potentially have led to different goal diagrams are lost. For instance, a group of stakeholders specifying a goal model for a user interface may decide to reduce softgoals ``easy to use'' and ``fast'' to one softgoal ``easy to use''. Thus, the resulting goal model will merely contain the softgoal ``easy to use'', but the discussion as well as the decision to reject ``fast'' are lost. This leads to a poor understanding of the problem and solution domain. In fact, empirical data suggest that this is an important reason of RE project failure~\cite{curtis1988field}.

Fourth and lastly, in goal models in general, ``rationale'' behind any decision is static and does not immediately impact the goal models when they change. That is, current goal modeling languages have limited support for reasoning about changing beliefs and opinions, and their effect on the goal model. A stakeholder may change his or her opinion, but it is not always directly clear what its effect is on the goal model. Similarly, a part of the goal model may change, but it is not possible to formally reason about the consistency of this new goal model with the underlying beliefs and arguments in the goal modeling languages. This becomes more problematic if the participants constructing the goal model change, since modeling decisions made by one group of stakeholders may conflict with the underlying beliefs put forward by another group of stakeholders.

\subsection{Research Questions}

The aim of this research is to resolve the above issues by developing a framework with a tool-support, which combines goal modeling approaches with argumentation approaches techniques from Artificial Intelligence (AI) research\cite{atkinson2007}. We have identified five important requirements for our framework: 
\begin{enumerate}
\item The argumentation techniques should be close to actual discussions stakeholders or designers have in the early requirements engineering phase.
\item 
The framework must have formal traceability links between elements of the goal model and underlying arguments.
\item 
Using these traceability links, it must be possible to compute the effect of changes in the goal model on the underlying arguments, and visa versa.
\item 
The framework must have tool support.
\item 
A methodology should be identified that allows the framework to be used by practitioners.
\end{enumerate}

In this context, the main research question is:

\begin{quote}
\textbf{RQ.} What are the constructs, mechanisms and rules for developing a framework that formally captures the discussions between stakeholders such that it can generate goal models?
\end{quote}

The five requirements above are the success criteria of our approach, that is, the satisfaction of the six requirements will result in a positive answer to the research question.

The main contribution of this article is the RationalGRL framework,  which combines the Goal-oriented Requirements Language (GRL) with a technique from argumentation and discourse modeling called \emph{argument schemes}~\cite{walton-etal2004}. We develop a metamodel specifying how our framework extends GRL, and we use this as the basis for a prototype web-based implementation~\footnote{The tool is available at <GITHUB LINK>}. We develop an initial list of argument schemes and critical questions by analyzing a set of transcripts containing discussion about an information system. Our framework is fully extensible, meaning that argument schemes and critical questions can be added by users for specific problem domains. We develop a methodology for using RationalGRL, which consists developing goal models and posing arguments in an integrated way. This methodology is intended to be used by practitioners in the field. We formalize RationalGRL using propositional logic, and develop algorithms for adding goal elements and arguments.

@@

This article is organized as follows. Section~\ref{sect:background} contains background and introduces our running example, the Goal-oriented Requirements Language (GRL)~\cite{}, and argument schemes. Section~\ref{sect:overview} provides a brief and high-level overview of our framework, together with a metamodel and the methodology. Section~\ref{sect:gmas} contains an in depth explanation of how we obtained an initial set of argument schemes by annotating transcripts from discussions about an information system, and in Section~\ref{sect:examples} we provide several examples of these arguments schemes. In Section~\ref{sect:formalframework} we formalize RationalGRL and our arguments schemes using propositional logic and we use well-known argumentation semantics to compute which arguments are accepted and which are rejected. We also develop various algorithms for the argument schemes in this section. In Section~\ref{sect:tool} we provide a brief overview of the prototype tool we developed for RationalGRL. Finally, Section~\ref{sect:discussion} contains a discussion, covering related work, future work, and a conclusion.

### 2017-06-27 overview.tex
many improvements (see email)

\section{RationalGRL overview and metamodel}
\label{sect:preview}

\todo{Marc}{all}{I suggest the following subsections, so the methodology becomes part of this section. What do you think of this?}

\subsection{RationalGRL overview}

\todo{Marc}{all}{Explain general idea of RationalGRL, perhaps by explaining the difference between Figure 2 and 5.}

@@

Extend the GRL metamodel with our concepts.

\subsection{RationalGRL methodology}

\todo{Marc}{Sepideh}{Finish this}

The methodology we propose in this paper is visualized in figure~\ref{fig:rationalgrl-methodology}. There are two main activities, depicted in grey, namely "goal modeling" and "argumentation". These are two separate activities that are being done in parallel.

\begin{figure}[ht]
\centering
\includegraphics[scale=0.4]{img/methodology}
\caption{The RationalGRL Methodology}
\label{fig:rationalgrl-methodology}
\end{figure}

>>>>

\section{RationalGRL Methodology and Metamodel}
\label{sect:overview}

In this section we present a high-level overview of the RationalGRL framework. In the first subsection we present a methodology, specifying how practitioners can use RationalGRL to create goal models with traceability links to underlying arguments. In the second subsection we link the new argumentation elements and relations to the existing elements and relations of GRL in a metamodel. This metamodel serves as the specification for an implementation. In Section~\ref{sect:tool} we briefly discuss our prototype implementation based on this metamodel.

\subsection{RationalGRL Methodology}

As mentioned earlier, the RationalGRL framework aims at integrating GRL models with the underlying discussions and arguments of the stakeholders during the analysis phase by combining PRAS with GRL. Our argumentation framework is developed based on the analysis of the CQs, practical reasoning and discussions of the stakeholders of the system. Initial GRL models are first created based on the analysis of non-functional requirements in the requirements specification documents, discussions with stakeholders and by refining the high-level goals into operationalized tasks.

In RationalGRL, both GRL and argumentation models are done iteratively. Argumentation models impact GRL models and result in refinement and changes in GRL. For example, an answer to the CQ \emph{Is the task possible?} can lead to rejection or acceptance of the task in GRL. The analysis of GRL models may result in changes in the underlying argumentation framework as well. For example, adding a new IE to the GRL model can result in posing new CQs with respect to the IE and its relationships. Figure~\ref{fig:rationalgrl-framework} shows RationalGRL and the relationships between the different parts. On the right-hand side of the framework, we have GRL models and on the left-hand side we have argumentation framework. The links between the two sides illustrates the impacts each side has on the other side and the changes that occur due to the impacts and the refinements.

\begin{figure}[ht]
\centering
\includegraphics[scale=0.4]{img/framework}
\caption{The RationalGRL Framework}
\label{fig:rationalgrl-framework}
\end{figure}

To develop a RationalGRL model, we follow the methodology shown  in Figure~\ref{fig:rationalgrl-methodology}. We already have an initial GRL model. The steps of the methodology are as follows:

\begin{figure*}[ht]
\centering
\includegraphics[scale=0.4]{img/methodology}
\caption{The RationalGRL Methodology}
\label{fig:rationalgrl-methodology}
\end{figure*}

\textbf{Instantiate Argument Schemes (AS)} -- In this step, we instantiate one of the AS from the argumentation framework. In the section we develop 13 arguments schemes which we use in our analysis. An example of an argument scheme is "Goal \emph{G} contributes to softgoal \emph{S}". An instantiation of an argument scheme thus corresponds to an argument for or against part of a goal model.

\textbf{Answer Critical Questions (CQs)} -- Instantiated arguments can be attacked by counter-arguments. CQs are ways in which AS can be attacked. Each argument scheme includes one or more CQs. For example, for the argument scheme mentioned above, we have two CQs as: \emph{Does the goal contributes to the softgoal?} and \emph{Does the goal contributes to some other softgoals?}. Note that, the answer to CQ can also result into ``conflict'' situation which we do not consider here. Answering a CQ may result in an instantiation of a new AS. Thus, it is possible to go back and forth between this step and the previous one.

\textbf{Decide on IE and the Relationships} -- The answers to the CQs can result in one of the four cases which impact the arguments and corresponding GRL IE: \textsf{INTRO}, \textsf{DISABLE}, \textsf{REPLACE} or \textsf{ATTACK}.  \textsf{INTRO} means that the argument scheme of the CQ does not get attacked and instead it creates a new argument. \textsf{DISABLE} means that the IE or the relationship related to the AS needs to be disabled or removed from the models. \textsf{REPLACE} introduces a new argument and attack the original argument at the same time. It basically replace the original element of the AS with a new one.  \textsf{ATTACK} is a generic counterargument which attacks any argument with another argument when a new evidence occur.

\textbf{Modify GRL Models} -- Based on the result of step three, the GRL models can be modified. In this case, either a new IE or a new relationships is introduced, an existing IE or relationship gets disabled (removed) from the model or finally an existing IE or relationships is replaced by a new one. This results in a new modified GRL. The new GRL model can then impact the argument schemes and instantiate another argument scheme.

In the next section, we provide a short example to illustrate the relationship between argumentation framework and goal models.

\subsubsection*{Example}
\label{sect:methodology-example}

Our examples comes from a transcript containing discussions about the development of an information system, and they are created as part of two master theses on improving design reasoning~\cite{masterthesis1,masterthesis2}. We provide several more examples in Section~\ref{sect:examples}.

A group of stakeholders is developing a goal model for a traffic simulator example and they are modeling the goal \emph{simulate} of the system using the RationalGRL methodology. This proceeds in the following way:
\begin{itemize}
\item
First they start at step \emph{Modify GRL models} (Figure~\ref{fig:rationalgrl-methodology}), and add the IE \texttt{Simulate} to GRL model (Figure~\ref{fig:examples:decomposition}, GRL model, top IE).
\item Next, they switch to the argumentation part (step \emph{Instantiate arguments schemes}). They answer CQ \emph{Does Simulate AND-decompose into any tasks?} with \emph{Yes: namely tasks ``Dynamic simulation'' and ``Static simulation}.
\item As a result, two argument schemes are created, namely:
\begin{itemize}
\item Actor \emph{System} has task \emph{Dynamic simulation}
\item Actor \emph{System} has task \emph{Static simulation}
\end{itemize}
\item In the GRL model, this corresponds to the addition of two tasks, and an AND-decomposition from the goal \emph{Simulate} to these two tasks.
\item Next, the stakeholders test the validity of their goal model by answering more CQ. They answer two CQs:
\begin{itemize}
\item
CQ \emph{Is task ``Dynamic simulation'' relevant} is answered with ``No, it is not relevant since the problem specification explicitly states dynamic simulations are not required''. Thus, the corresponding GRL IE is disabled.
\item The decomposition is changed from AND to OR, since it turned out not both tasks can be implemented together.
\end{itemize}
\end{itemize}

The resulting RationalGRL model is shown in Figure~\ref{fig:examples:decomposition}, which can be found in Section~\ref{sect:examples}, where we will also explain the visualization of the examples in more detail.

@@

\todo{Marc}{all}{This section is copy/pasted from RENext, we should revise the metamodel and the text.}

Figure~\ref{fig:metamodel} depicts the metamodel linking the main elements of our argumentation extension to the main GRL elements. The part below the dashed horizontal line depicts GRL elements. A \textsf{GRL Diagram} (bottom) contains zero or more \textsf{IntentionalElements}, either a \textsf{Goal}, a \textsf{Softgoal}, a \textsf{Task}, or a \textsf{Belief}. An \textsf{ElementLink} is either a \textsf{Contribution}, a \textsf{Decomposition}, or a \textsf{Dependency} and contains an \textsf{IntentionalElements} as source and as target.

The part above the horizontal dashed line depicts the concepts we introduced in the previous section. The top-left element \textsf{Argument} (see Section~\ref{sect:arguments}) can attack other arguments (Section~\ref{sect:attacks}) and generalizes both a \textsf{Formula} and an \textsf{Inference}. That is, both these elements can be arguments. A formula has an \textsf{AcceptStatus} (Section~\ref{sect:extensions}). An \textsf{Inference} is from a set of \textsf{Arguments} as premise and a \textsf{Formula} as conclusion. The \textsf{InferenceType} is either strict or defeasible (Section~\ref{sect:aspic}). A \textsf{Formula} is either a \textsf{Modality} (modal formula), a \textsf{Proposition}, a \textsf{BinaryOperation}, or a \textsf{Negation} (negated proposition). The \textsf{Modality} can be either B (belief), G (goal), E (evidence), or A (action). A \textsf{BinaryOperation} is either a \textsf{Disjunction}, a \textsf{Conjunction}, or a \emph{contributes\_to} \textsf{Connective}~(\ref{sect:aspic}).

Finally, the red arrows depict how the two metamodels are integrated. The left arrow connects a GRL \textsf{IntentionalElement} with an argumentation \textsf{Modality}, where the mapping is denoted with red text. Thus, an intention element traces to an argument, which is always a modal formula. The right arrow connects the GRL \textsf{ElementLink} with the argumentation \textsf{Connective} (i.e. a \emph{contributes\_to} connective). Thus, an arrow between elements of a GRL model corresponds to \emph{contributes\_to} connectives in the argumentation framework.

\begin{figure}[h!]
\includegraphics[scale=0.61]{img/metamodelimg}
\caption{The Metamodel \todo{Marc}{all}{This has to be revised.}}
\label{fig:metamodel}
\end{figure}
